Changes in gene expression associated with transitions in reproductive meristem identity during early development of rice inflorescence.

### 2015-12-10 figures.Rmd
Added NEWS for emailing updates Fixed figure 4 (alog) (fixes #21) More SPL discussion (fixes #17)

---
title: "Figures"
csl: csl/the-plant-journal.csl
header-includes:
  \usepackage{graphicx}
  \usepackage{caption}
  \usepackage{fixltx2e}
  \usepackage{rotating}
  \renewcommand{\thefigure}{S\arabic{figure}}
  \renewcommand{\thetable}{S\arabic{table}}
documentclass: tufte-handout
classoption: a4paper
output:
  rmarkdown::tufte_handout:
    dev: CairoPDF
    keep_tex: yes
    fig_crop: no
bibliography: bib/papersLib.bibtex
---

```{r Setup, include=FALSE, results="hide", warning=FALSE}
knitr::opts_chunk$set(fig.path='figures/',
               cache.path='cache/',
               fig.align = 'center', 
               dpi=600,
               fig.show='hold',
               fig.width=3.150,
               fig.height=8.661,
               echo=FALSE,
               results="hide",
               message=FALSE,
               warning=FALSE)

@@

```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 5/2, fig.cap = "Expression of selected cytokinin-related genes (\\textbf{a}), detected \\emph{G1}-Like (G1L/ALOG) genes (\\textbf{b}) and \\emph{STMADS11}-like MADS genes (\\textbf{c}) in inflorescence meristems in transcripts per million (TPM). A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected. \\emph{GN1A} (\\emph{CKX2}) was not detected in the LMD dataset, but \\emph{LOG1} is expressed in all meristem types, and two other genes possibly related to cytokinin activation, \\emph{LABA1} (\\emph{LOGL6}) and \\emph{LOGL1}, are expressed more highly in ePBM/SBM and SM than in RM and PBM. Only three G1L genes (\\emph{G1L1}, \\emph{G1L2} and \\emph{G1L5}/\\emph{TAWAWA1}) were detected, but they share a similar pattern of expression. Two reported \\emph{STMADS11}-like (\\emph{SVP}-like) targets of G1L5, \\emph{MADS22} and \\emph{MADS55} [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other \\emph{STMADS11}-like gene, \\emph{MADS47}, was only detected in one library."}
# set layout
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(5, 2)))

# print the plots
print(f_alogFamily_a, vp = grid::viewport(layout.pos.row = 1:2, layout.pos.col = 1:2))
print(f_alogFamily_b, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 1))
print(f_alogFamily_c, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 2))
```

>>>>

---
title: "Figures"
date: "\\vspace{0 mm}"
csl: csl/the-plant-journal.csl
header-includes:
  \usepackage{graphicx}
  \usepackage{caption}
  \usepackage{fixltx2e}
  \usepackage{rotating}
  \renewcommand{\thefigure}{\arabic{figure}}
  \renewcommand{\thetable}{\arabic{table}}
documentclass: tufte-handout
classoption: a4paper
output:
  rmarkdown::tufte_handout:
    dev: CairoPDF
    fig_crop: no
bibliography: bib/papersLib.bibtex
---

```{r Setup, include=FALSE, results="hide", warning=FALSE}
## grep queries:
  # bold: \*\*(\w*)\*\* -> \\\\textbf{\1}
  # italic: \*(\w*)\* -> \\\\emph{\1}
knitr::opts_chunk$set(fig.path='figures/',
               cache.path='cache/',
               fig.align = 'center', 
               dpi=600,
               fig.show='hold',
               fig.width=3.150,
               fig.height=8.661,
               echo=FALSE,
               results="hide",
               message=FALSE,
               warning=FALSE)

@@

```{r f_alogFamily, fig.width = 6.614, fig.height = 6.614, fig.cap = paste0("Expression of selected \\emph{STMADS11}-like MADS genes (\\textbf{a}), detected \\emph{G1}-Like (G1L/ALOG) genes (\\textbf{b}), \\emph{SPL} genes and a microRNA precursor (\\textbf{c}) and cytokinin-related genes (\\textbf{d}) in inflorescence meristems in transcripts per million (TPM). A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected. Two reported \\emph{STMADS11}-like (\\emph{SVP}-like) targets of G1L5, \\emph{MADS22} and \\emph{MADS55} [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other \\emph{STMADS11}-like gene, \\emph{MADS47}, was only detected in one library. Only three G1L genes (\\emph{G1L1}, \\emph{G1L2} and \\emph{G1L5}/\\emph{TAWAWA1}) were detected, but they share a similar pattern of expression. Several \\emph{SPL} genes and a co-regulated microRNA precursor, \\emph{MIR319A}, recovered from cluster ", getClusterName('small', 'dec'), ", are highly expressed in RM and PBM before decreasing in expression in ePBM/SBM and SM. \\emph{GN1A} (\\emph{CKX2}) was not detected in the LMD dataset, but \\emph{LOG1} is expressed in all meristem types, and two other genes possibly related to cytokinin activation, \\emph{LABA1} (\\emph{LOGL6}) and \\emph{LOGL1}, are expressed more highly in ePBM/SBM and SM than in RM and PBM.")}
# set layout
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(4, 4)))

# print the plots
print(f_alogFamily_a, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 1))
print(f_alogFamily_b, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 2))
print(f_alogFamily_c, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 3:4))
print(f_alogFamily_d, vp = grid::viewport(layout.pos.row = 4, layout.pos.col = 1:4))
```

### 2015-12-10 ms.Rmd
Added NEWS for emailing updates Fixed figure 4 (alog) (fixes #21) More SPL discussion (fixes #17)

@@

Overexpression of the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff]. Yeast two-hybrid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff]. *MADS22* and *MADS55* also decrease in expression between PBM and SM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c). The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes. Of the G1L genes, which are annotated with the Pfam domain PF04852, only three were detected in the dataset, with two of them present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *G1L5* respectively), and the third, *G1L1*, not present in the clustering results but having a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b). *g1l5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNA interference [@Yoshida:2013ff]. This may suggest functional redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with more than one being affected by the RNAi construct used against *G1L5*. Although the increase in secondary branches linked to *G1L5* suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *g1l5* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *G1L5*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem that remains to be reported.

@@

The strong enrichment of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the PBM and their presence in cluster `r getClusterName('small', 'dec')` suggests involvement in the early stages of inflorescence development. *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM, suggesting a redundant function in indeterminate axillary meristems. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc]. Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [@Lu:2013hy], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O. sativa* ssp. *japonica* varieties. Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf]. WFP also binds to the promoters of *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].

A homolog of *LOG1*, *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*), is present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM. In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by @Han:2014dt]. The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM, small panicles with small SAMs and branching defects [@Kurakawa:2007go]. Similarly, reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield *O. sativa* ssp. *indica* varieties [@Ashikari:2005eg]. Although *GN1A* transcripts were not detected in this LMD dataset (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a), these results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*) gene provides evidence for an inverse role. The LABA1 protein present in *O. rufipogon*, which has an identical primary sequence to the protein from *O. sativa* ssp. *japonica* cv. Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O. rufipogon* into an *O. sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [@Hua:2015bs], but reduces the number of grains per panicle and the number of tillers per plant [@Gu:2015hy]. *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs]. *LABA1* was detected in the ePBM/SBM and SM samples but not the RM or PBM samples. Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation along the course of axillary meristem determination (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). Such a mechanism would support a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level. A type-A *RESPONSE REGULATOR* (*RR3*) is present in cluster `r getClusterName('dec', 'small')` (containing genes more highly expressed in RM), highlighting the possible complexity of the roles of cytokinins in inflorescence development.

@@

```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 5/2}
#grid::grid.draw(f_alogFamily)

# set layout
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(5, 2)))

# print the plots
print(f_alogFamily_a, vp = grid::viewport(layout.pos.row = 1:2, layout.pos.col = 1:2))
print(f_alogFamily_b, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 1))
print(f_alogFamily_c, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 2))
```

**`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**. Expression of selected cytokinin-related genes (**a**), detected *G1*-Like (G1L/ALOG) genes (**b**) and *STMADS11*-like MADS genes (**c**) in inflorescence meristems in transcripts per million (TPM). A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected. *GN1A* (*CKX2*) was not detected in the LMD dataset, but *LOG1* is expressed in all meristem types, and two other genes possibly related to cytokinin activation, *LABA1* (*LOGL6*) and *LOGL1*, are expressed more highly in ePBM/SBM and SM than in RM and PBM. Only three G1L genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA1*) were detected, but they share a similar pattern of expression. Two reported *STMADS11*-like (*SVP*-like) targets of G1L5, *MADS22* and *MADS55* [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other *STMADS11*-like gene, *MADS47*, was only detected in one library.

>>>>

# max width (2 col): 6.614
# max width (1 col): 3.150
# max height : 8.661

@@

Overexpression of the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff]. Yeast two-hybrid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff]. *MADS22* and *MADS55* also decrease in expression between PBM and SM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes. Of the G1L genes, which are annotated with the Pfam domain PF04852, only three were detected in the dataset, with two of them present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *G1L5* respectively), and the third, *G1L1*, not present in the clustering results but having a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b). *g1l5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNA interference [@Yoshida:2013ff]. This may suggest functional redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with more than one being affected by the RNAi construct used against *G1L5*. Although the increase in secondary branches linked to *G1L5* suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *g1l5* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *G1L5*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem that remains to be reported.

@@

The strong enrichment of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the PBM and their presence in cluster `r getClusterName('small', 'dec')` suggests involvement in the early stages of inflorescence development. *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c), suggesting a redundant function in indeterminate axillary meristems. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx], *SPL* gene expression is controlled by *miR156* and *miR529* [**ref Jeong 2011**; @Wang:2015jx]. The precursor for another microRNA, *MIR319A*, was recovered from the same cluster as *SPL3* and *SPL12* and has a similar expression pattern to these genes, *SPL17* and *WFP* (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c). Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [@Lu:2013hy], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O. sativa* ssp. *japonica* varieties. Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf]. WFP also binds to the promoters of *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].

A homolog of *LOG1*, *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*), is present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM. In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by @Han:2014dt]. The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM, small panicles with small SAMs and branching defects [@Kurakawa:2007go]. Similarly, reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield *O. sativa* ssp. *indica* varieties [@Ashikari:2005eg]. Although *GN1A* transcripts were not detected in this LMD dataset (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`d), these results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*) gene provides evidence for an inverse role. The LABA1 protein present in *O. rufipogon*, which has an identical primary sequence to the protein from *O. sativa* ssp. *japonica* cv. Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O. rufipogon* into an *O. sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [@Hua:2015bs], but reduces the number of grains per panicle and the number of tillers per plant [@Gu:2015hy]. *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs]. *LABA1* was detected in the ePBM/SBM and SM samples but not the RM or PBM samples. Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation along the course of axillary meristem determination (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). Such a mechanism would support a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level. A type-A *RESPONSE REGULATOR* (*RR3*) is present in cluster `r getClusterName('dec', 'small')` (containing genes more highly expressed in RM), highlighting the possible complexity of the roles of cytokinins in inflorescence development.

@@

```{r f_alogFamily, fig.width = 6.614, fig.height = 6.614}
# set layout
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(4, 4)))

# print the plots
print(f_alogFamily_a, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 1))
print(f_alogFamily_b, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 2))
print(f_alogFamily_c, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 3:4))
print(f_alogFamily_d, vp = grid::viewport(layout.pos.row = 4, layout.pos.col = 1:4))
```

**`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**. Expression of selected *STMADS11*-like MADS genes (**a**), detected *G1*-Like (G1L/ALOG) genes (**b**), *SPL* genes and a microRNA precursor (**c**) and cytokinin-related genes (**d**) in inflorescence meristems in transcripts per million (TPM). A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected. Two reported *STMADS11*-like (*SVP*-like) targets of G1L5, *MADS22* and *MADS55* [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other *STMADS11*-like gene, *MADS47*, was only detected in one library. Only three G1L genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA1*) were detected, but they share a similar pattern of expression. Several *SPL* genes and a co-regulated microRNA precursor, *MIR319A*, recovered from cluster `r getClusterName('small', 'dec')`, are highly expressed in RM and PBM before decreasing in expression in ePBM/SBM and SM. *GN1A* (*CKX2*) was not detected in the LMD dataset, but *LOG1* is expressed in all meristem types, and two other genes possibly related to cytokinin activation, *LABA1* (*LOGL6*) and *LOGL1*, are expressed more highly in ePBM/SBM and SM than in RM and PBM.
